---
title: "Exercise 3 and 4"
author: "Dan Ovando"
date: "September 27, 2015"
output: html_document
---



```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)

```

```{r set things up}

rm(list = ls())
set.seed(448)
library(gridExtra,quietly = T)
library(ggplot2, quietly = T)
library(plyr, quietly = T)
library(dplyr, quietly = T)
library(tidyr, quietly = T)
library(knitr, quietly = T)

```

# Sample Importance Resampling - Example 3 with Grey Whales

The problem here is that we don't believe the standard errors 

```{r SIR example}


# =================================================================================

DoSir <- function(Nout=1000,Model,DropYears = NA)
{
  
  # Read in the basic data
  TheData <- ReadData(DropYears)
  Yr1 <- TheData$CatchYr[1]
  Catch <- TheData$CatchVal
  SurveyEst <- TheData$SurveyEst
  SurveyCV <- TheData$SurveyCV
  SurveyYr <- TheData$SurveyYr
  Nyears <- length(Catch)
  years <- TheData$CatchYr
  time  <- length(years)
  # Storage for the parameters and the final depletion
  #   Vals <- as.data.frame(matrix(0,ncol=6,nrow=Nout))
  
  Vals <- list()
  
  # Storage for the total likelihood encountered in the first stage sampling
  # and the number of draws
  AveLike <- 0
  Ntest <- 0
  
  # Reset parameters for SIR
  Threshold <- exp(0)
  Cumu <- 0
  Ndone <- 0
  
  output <- (matrix(NA,nrow = (time),ncol = 1))
  colnames(output) <- 'n'
  while (Ndone < Nout)
  {
    # Generate from priors
    
    r <- runif(1,0,.15)
    
    Pop1965 <- runif(1,10000,15000)
    
    AddCV <- runif(1,.1,.2)
    
    K <- runif(1,20000,50000)
    
    # Call the population model
    Pop <- PopModel(Catch = Catch,r = r,K = K,years = years,InitPop = Pop1965,ExponModel = Model,output = output)
    
    #     survey <- data.frame(TheData$SurveyYr,TheData$SurveyEst,TheData$SurveyCV)
    # 
    #     colnames(survey) <- c('year','SurveyEst','SurveyCV')
    # 
    #     Pop <- join(Pop,survey, by = 'year')
    
    #     ggplot(Pop,aes(year,n)) + geom_point() + geom_line(aes(year,SurveyEst))
    #
    
    #write the pop model
    # Compute the negative log-likelihood and hence the likelihood
    NegLogLike <- Likelihood(Pop = Pop,SurveyYr-Yr1+1,SurveyEst,SurveyCV,AddCV)
    TheLike <- exp(-1*NegLogLike-32.19)
    
    # Determine if a parameter vector is to be saved
    Cumu <- Cumu + TheLike
    
    AveLike <- AveLike + TheLike
    Ntest <- Ntest +1
    
    while (Cumu > Threshold & Ndone < Nout)
    {
      Ndone <- Ndone + 1
      show(Ndone)
      Cumu <- Cumu - Threshold
      Vals[[Ndone]] <- data.frame(Ndone,years,K,r,Pop1965,AddCV,NegLogLike,Pop,(Pop/K))
      
    }
  }
  
  Vals <- ldply(Vals)
  
  colnames(Vals) <- c('iteration','year','K','r','Pop1965','AddCV','NegLogLike','Pop','NvK')
  
  Vals$AveLike <- AveLike/Ntest
  
  return(Vals)
}

# =================================================================================

Likelihood <- function(Pop,SurveyYr,SurveyEst,SurveyCV,AddCV)
{
  # Account for the additional CV
  UseCV <- sqrt(SurveyCV^2+AddCV^2)
  
  # Extract the predictions corresponding to the observations and compute the negatuve log-likelihood
  Preds <- Pop[SurveyYr]
  Residuals <- log(UseCV)+0.5*(log(Preds)-log(SurveyEst))^2/UseCV^2
  LogLike <- sum(Residuals)
}

# =================================================================================
PopModel <- function(Catch,r,K,years,InitPop,ExponModel,output)
{
  
  time <- length(years)
  
  #   colnames(output) <- c('year','catch','n')
  
  #   output$catch <- Catch
  
  output[1,1] <- InitPop
  
  #   output$year <- years
  
  if (ExponModel == 'ExponModel')
  {
    
    for (t in 2:time)
    {
      
      output[t,1] <- (1+r)*output[t-1,1] - Catch[t-1]
      #             output$n[t] <- pmax(1e-5,(1+r)*output$n[t-1] - output$catch[t-1])
      
      
    }
    
  }
  if (ExponModel == 'Schaefer')
  {
    
    for (t in 2:time)
    {
      output[t,1] <- pmax(1e-5,output[t-1,1] + (output[t-1,1]*r)*(1-output[t-1]/K) - Catch[t-1])
    }
    
  }
  
  return(output)
}

# =================================================================================
ReadData <- function(DropYears)
{
  lecture <- 'Workshop Day 3'
  TheData1 <- read.csv(paste('Fish 558 Workshop/',lecture,'/Ex3a.csv', sep = ''),header=TRUE, stringsAsFactors = F)
  
  colnames(TheData1) <- c('year','pop','cv')
  
  TheData2 <- read.csv(paste('Fish 558 Workshop/',lecture,'/Ex3b.csv', sep = ''),header=TRUE, stringsAsFactors = F)
  
  colnames(TheData2) <- c('year','thing')
  
  if (class(DropYears) == 'numeric')
  {
    TheData1 <- subset(TheData1,!(year %in% DropYears))
#     TheData2 <- subset(TheData2,!(year %in% DropYears))
    
  }
  
  
  Outs <- NULL
  Outs$SurveyYr <- TheData1[,1]
  Outs$SurveyEst <- TheData1[,2]
  Outs$SurveyCV <- TheData1[,3]
  Outs$CatchYr <- TheData2[,1]
  Outs$CatchVal <- TheData2[,2]
  return(Outs)
}

# =================================================================================
set.seed(443)

ExpoModel <- DoSir(Nout=200,Model= 'ExponModel')

SchModel <- DoSir(Nout=200,Model='Schaefer')

SchModel_drop <- DoSir(Nout=200,Model='Schaefer', DropYears = c(2001,2002))

ExpoModel_drop <- DoSir(Nout=200,Model='ExponModel', DropYears = c(2001,2002))


ggplot(subset(SchModel, year == max(year)),aes(x = K,y = r)) + geom_point(aes(fill = NegLogLike), shape = 21, size = 2) + scale_fill_gradient(low = 'green',high = 'red')


SchModel$NearK <- SchModel$NvK >= 0.9

prob_recovered <- mean(SchModel$NearK[SchModel$year == max(SchModel$year)], na.rm = T)

```

So, the bayes factor for the schaefer model is `r mean(SchModel$AveLike)/mean(ExpoModel$AveLike)`. This means that the probability of the *Schaefer* model is `r mean(SchModel$AveLike)/(mean(SchModel$AveLike) + mean(ExpoModel$AveLike))` while the probability of the *Exponential* model is `r mean(ExpoModel$AveLike)/(mean(SchModel$AveLike) + mean(ExpoModel$AveLike))` . 

The bayes factor for the schaefer model without 2001 2002 relative to the exponential model is `r mean(SchModel_drop$AveLike)/mean(ExpoModel_drop$AveLike)`.while the probability of the *Schaefer* model with less years is `r mean(SchModel_drop$AveLike)/(mean(ExpoModel_drop$AveLike) + mean(SchModel_drop$AveLike))` . 

The probability of the popuataltion begin >= 90% of *K* in `r max(Year)` is `r 100*prob_recovered`%
