---
title: "Untitled"
author: "Dan Ovando"
date: "September 26, 2015"
output: ioslides_presentation
---

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r global_options, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(fig.path='Figs/', echo=FALSE, warning=FALSE, message=FALSE)

library(gridExtra)
library(ggplot2)
library(plyr)
library(dplyr)
library(tidyr)
library(knitr)
```

```{r}
devtools::load_all('~/R Functions/mletools')

devtools::load_all('~/R Functions/PopulationModels')

hakedata <- read.csv('namibian_hake_data.csv', stringsAsFactors = F)

kable(summary(hakedata), caption = 'Summary of Namibian hake data')

hake_plot <- (ggplot(gather(hakedata,'Data','Value',2:3),aes(year,Value, fill = Data)) + geom_point(shape = 21) + facet_grid(Data~., scales = 'free_y') +scale_fill_discrete(guide = FALSE))

r_guess <- 0.1

k_guess <- 10*max(hakedata$catch)

q_guess <- .001

sdev_guess <- sd(hakedata$cpue)

schaefer_hake_fit <- nlminb(start = c(r_guess,k_guess,q_guess,sdev_guess), objective = schaefer_mle,dat = hakedata,
                            use = 'optimization', pars_to_opt = c('r','k','q','sdev'), lower = c(0.01,.01,1e-7,1e-7),
                            upper = c(1,Inf,1,Inf))


schaefer_fit <- schaefer_mle(params =schaefer_hake_fit$par, dat = hakedata)

```

## Summary of Data

```{r, echo=T}

a <- 3 + 4

hake_plot
```

## Slide with Plot

```{r, echo=FALSE}
sch_cpue_fit <- (ggplot(schaefer_fit$dat,aes(year,cpue, shape = 'Observed')) + 
                   geom_point() + 
                   geom_line(aes(year,predicted_cpue,linetype = 'Predicted'), color = 'red') + theme(legend.title = element_blank()))

sch_cpue_fit
```

