---
title: "FISH-558-Shiny-Homework-5"
author: "Dan Ovando"
date: "November 13, 2015"
output: html_document
runtime: shiny
---


The goal of this assignment is to provide and MSE/decision analysis around the management of a mythical endangered species. There is uncertainty around parameters and a variety of potential management actions that need considering. 

```{r load parameters}

# n_at_age <- matrix(NA,nrow = sim_years, ncol = ages)

```

## Task A

Find the relationship between the mean and CB of a beta distribtuion and the shape1 and shape 2 parameters for rbeta. The goal here is to take the mean and CV parameters for the beta distribution provided by Andre and convert them into R usable form. 

The beta distribution is a continuous probability distribution defined on the interval [0,1], that can take a wide array of shapes 

```{r, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```




